---
title: "Introduction to R--Exercise Solutions"
author: "Chris Hansen"
date: "16/10/2016"
output:
  html_document:
    self_contained: yes
    toc: yes
    toc_float: yes
  html_notebook:
    code_folding: show
    self_contained: yes
    toc: yes
    toc_float: yes
---

<style type=text/css>
.table th:not([align]){
  text-align: center;
}
img{
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(comment = NA, fig.align = "center")
```


## Basic Types and Operations

### Assignment

### Binary operators

### Logical operators {.smaller}


## Data Types and Structures

### Numeric

### Character (strings)

#### Exercises

1. Store the string `"Hello, World!"` as a variable called `s`.
```{r}
s <- "Hello, World!"
```
2. Split `s` programmatically into the two words `Hello` and `World!` (hint:
`?strsplit`).
```{r}
strsplit(s, ", ")[[1]]
```
3. Replace `World` with your own name (hint: `?gsub`).
```{r}
gsub("World", "Chris", s)
```
4. Replace `World` with your own name (hint: `?substr`).
```{r}
substr(s, 8, 12)
substr(s, 8, 12) <- "Chris"
s
```
5. Find the string containing `needle` in the vector
`c("haystack needle", "needle")` (hint: `?agrep`).
```{r}
# when value=TRUE, return the matching strings
agrep("needle", c("haystack needle", "haystack"), value=TRUE)
# when value=FALSE, return index of the matching strings
haystack <- c("haystack needle", "haystack")
haystack[agrep("needle", haystack, value=FALSE)]
```
6. Find the string containing `needle` in the vector
`c("haystack needle", "needle")` (hint: `?grep`).
```{r}
# usual wild cards
grep("*needle*", c("haystack needle", "haystack"), value=TRUE)
# PERL regular expression
grep("^.*(needle).*?", c("haystack needle", "haystack"), 
     perl=TRUE, value=TRUE)
```

### Boolean / logical values

### `NULL` 

### `NA`

### Vectors

#### Exercises

1.  Store all the odd numbers less than 20 as `x`.
```{r}
(x <- seq(from=1, to=20, by=2))
```
2.  Get the first 5 elements of `x` (hint: `?head`).
```{r}
head(x, n=5)
# or 
x[1:5]
```
3.  Get the last 5 elements of `x` (hint: `?tail`).
```{r}
tail(x, n=5)
# or
x[(length(x)-4):length(x)]
```
4.  Get `x` in reverse order (hint: `?rev`).
```{r}
rev(x)
# or
x[length(x):1]
```
5.  Create a vector containing 5 random numbers distributed $U(0,1)$ and store it
    as `y` (hint: `?runif`).
```{r}
set.seed(1)
(y <- runif(5, min = 0, max = 1))
```
6.  Sort `y` (hint: `?order`).
```{r}
y[order(y)]
```
7.  Sort `y` (hint: `?sort`).
```{r}
sort(y)
```
8.  Calculate the mean of `y` from first principles (hint: `?sum`, `?length`).
```{r}
sum(y)/length(y)
```
9.  Calculate the mean of `y` (hint: `?mean`).
```{r}
mean(y)
```
10. Calculate the standard deviation of `y` (hint: `?sd`).
```{r}
sd(y)
```
11. Create a new vector, `z`, containing 100 numbers distributed $N(0,1)$ (hint:
    `?rnorm`).
```{r}
set.seed(1)
z <- rnorm(n=1000, mean=0, sd=1)
```
12. Look at the distribution of `z` (hint: `?cut`, `?table`).
```{r}
table(cut(z, breaks=5))
```
13. Look at the distribution of `z` (hint: `?hist`).
```{r}
hist(z, col="lightgrey", border="white", freq=FALSE)
lines(density(z))
```


### Matrices

#### Exercises 

1. Use the ```solve``` function to solve the following system of equations:
\[
\begin{aligned}
-2x + y &= 1\\
  x + y &= 4
\end{aligned}
\]
```{r}
A <- matrix(c(-2,1,1,1), nrow=2)
b <- matrix(c(1,4))
solve(A,b)
```
2. Plot the two lines, to show the solution visually (hint: `?curve`, `points`).
```{r}
curve(1 + 2*x, from=0, to=2, xlab="x", ylab="y")
curve(4 - x, from=0, to=2, add=TRUE, col="grey")
points(1, 3, pch=16, col="red")
```


### Arrays

### Lists


## Data frames

### Importing / exporting data

### Basics

### Subsetting

#### Exercises

1. R has a good number of built-in datasets.  Some are already loaded as part of
   the base package, others have to be explicitly loaded with the `data`
   function.  What happens if you call `data()` without any arguments, though?
2. One of the more famous datasets, examine the basic structure of the `iris`
   dataset.  List the number of rows, number of columns, and the column names.
```{r}
str(iris) # look at the structure
nrow(iris) # number of rows
ncol(iris) # number of columns
colnames(iris) # column names
```
3. How many different species of flower are in the iris data (hint: `?unique`,
   `?table`)?
```{r}
length(unique(iris$Species))
#or
length(table(iris$Species))
```
4. If you're into linear regression, Anscombe's data is pretty famous.  What
   happens if you call `plot` on the `anscombe` dataset?
```{r}
plot(anscombe)
```

5. Summarise each of the variables in the ```anscombe``` dataset.
```{r}
summary(anscombe)
```
6. Using the `anscombe` data, fit the following linear models: `y1~x1`, `y2~x2`,
   `y3~x3`, and `y4~x4` (hint: ```?lm```).
```{r}
summary(lm(y1~x1, data=anscombe))
summary(lm(y2~x2, data=anscombe))
summary(lm(y3~x3, data=anscombe))
summary(lm(y4~x4, data=anscombe))
```


### Creating new variables

### Combining data 

### Summarising / aggregating data


## Functions

### Basics

### Parameters

### `...`

### Variable scope


## Program Flow

### `if`, `if else`

#### Exercises

1. Write a plot function that takes two parameters: `x`, a numeric vector; and
   `difference`, a boolean.  When `difference` is `TRUE`, plot the difference of
   `x`, otherwise plot `x`.  Do this with a single `if` statement.  Hint:
   `?diff`.
```{r}
myplot <- function(x, difference, ...){
   if (difference) x <- diff(x)
   plot(x, ...)
}

par(mfrow=c(1,2), mar=par()$mar-c(1,0,2,0), cex=0.8)

x <- rnorm(20, 100)

myplot(x, difference=FALSE, type="l")
myplot(x, difference=TRUE, type="l")
```
2. Write a plot function that takes two parameters: `x`, a numeric vector; and
   `difference`, a boolean.  When `difference` is `TRUE`, plot the difference
   of `x`, otherwise plot `x`.  Do this with an `if-else`.
```{r}
myplot <- function(x, difference, ...){
   if (difference) plot(diff(x), ...)
   else plot(x, ...)
}

par(mfrow=c(1,2), mar=par()$mar-c(1,0,2,0), cex=0.8)

x <- rnorm(20, 100)

myplot(x, difference=FALSE, type="l")
myplot(x, difference=TRUE, type="l")
```

### ```for```

### `while`


## Plotting with Base Graphics

### The `plot` function

### Specific plot types


## Plotting with `ggplot2`


## HTML and javascript

### `plotly`

### `highcharter` (highcharts)

### `leaflet`

### `DT` (Data Tables)


## Case Study: New Zealand Income Survey Super SURF

#### Exercises

```{r importsurf, cache=TRUE}
nzis <- read.csv("data/NZIS_Super_SURF.txt")
```

1.  Use `dplyr::group_by` and `dplyr::summarise` to calculate the mean of
    `weekly_income` for each replicate (i.e. for each unique `surf_ID`.
```{r, messsage = FALSE}
library(dplyr)

rep.means <- group_by(nzis, surf_ID) %>% 
   summarise(mean_weekly_income=mean(weekly_income))

head(rep.means, 5)
```
2.  Calculate the mean of the 100 mean weekly income estimates (hint: `?mean`).
```{r}
(mu <- mean(rep.means$mean_weekly_income))
```
3.  Calculate the variance of the 100 mean weekly income estimates (hint:
    `?var`).  This is an estimate of the sampling variance.
```{r}
(SampVar <- var(rep.means$mean_weekly_income))
```
4.  Calculate the standard error--just the square root of the sampling variance.
```{r}
(StdErr <- sqrt(SampVar))
```
5.  Calculate the SNZ sampling error--just 1.96 times the standard error (well, 
    actually, `qnorm(0.975)`=`r qnorm(0.975)` times the standard error).
```{r}
(SampErr <- 1.96 * StdErr)
```
6.  Calculate the relative sampling error (RSE)--just the sampling error divided
    by the mean estimate (times 100%).
```{r}
(RelSampErr <- SampErr / mu * 100)
```
7.  Make a histogram of the 100 mean estimates (hint: `?hist`).
```{r}
hist(rep.means$mean_weekly_income, freq=FALSE, xlab="mean ($)", col="grey", 
   border="white", main="mean weekly income for 100 replicates")
```
8.  Redraw the above histogram, but overlay a curve approximating the sampling
    distribution (hint: `?curve`).
```{r}
hist(rep.means$mean_weekly_income, freq=FALSE, xlab="mean ($)", col="grey", 
   border="white", main="mean weekly income for 100 replicates")
curve(dnorm(x, mu, StdErr), from=mu-3*StdErr, to=mu+3*StdErr, col="red", add=TRUE)
```
9.  The `nzis` dataset is effectively 100 datasets stacked on top of each other.
    Subset `nzis` to a single dataset, and store as `nzis.sub`.
```{r}
nzis.sub <- subset(nzis, surf_ID==1)
head(nzis.sub)
table(nzis.sub$surf_ID)
```
10. Make sure that `age_category` is categorical (hint: `?factor`).
```{r}
nzis.sub$age_category <- factor(nzis.sub$age_category, levels=seq(25,60,5),
   labels=c("25-29 years", "30-34 years", "35-39 years", "40-44 years",
            "45-49 years", "50-54 years", "55-59 years", "60-64 years"))
str(nzis.sub$age_category)
```
11. Make sure the `sex` variable is categorical, and assign levels such that 1
    is male and 0 is female.
```{r}
nzis.sub$sex <- factor(nzis.sub$sex, levels=c(1,0), labels=c("Male", "Female"))
str(nzis.sub$sex)
```
12. Make sure the `ethnicity` variable is categorical, and assign levels as
    follows: 1=European, 2=M&#257;ori, 3=Pacific, 4=Other, 5=M&#257;ori
    combination, and 6=non-M&#257;ori combination.
```{r}
nzis.sub$ethnicity <- factor(nzis.sub$ethnicity, levels=1:6, 
   labels=c("European", "M\u0101ori", "Pacific", "Other", "M\u0101ori combination", 
            "non-M\u0101ori combination"))
str(nzis.sub$ethnicity)
```
13. Make sure the `highest_qualification` variable is categorical, and assign
    levels as follows: 1=no qualification, 2=school, 3=vocational / trade,
    4=bachelor or higher, 5=other post-school.
```{r}
nzis.sub$highest_qualification <- factor(nzis.sub$highest_qualification, levels=1:5, 
   labels=c("no qualification", "school", "vocational / trade", "bachelor or higher",
            "other post-school"))
str(nzis.sub$ethnicity)
```
14. Plot `weekly_income` against `weekly_hours`--note that both variables are
    continuous.  Add / remove / change options to make the plot more attractive
    if you want.
```{r}
plot(nzis.sub$weekly_hours, nzis.sub$weekly_income, col=rgb(0,0,0,0.2), pch=16,
     xlab="weekly hours", ylab="weekly income ($)", main="")
```
15. Plot `weekly_income` against `highest_qualification`--note that
    `highest_qualification` is categorical (so a simple scatter probably isn't
    the best plot choice).
```{r}
boxplot(weekly_income ~ highest_qualification, data=nzis.sub)
```
16. Plot `weekly_income` against both `sex` and `ethnicity`.
```{r, warning = FALSE}
par(mar = par()$mar + c(0, 3, 0, 0), cex = 0.75)
boxplot(weekly_income ~ abbreviate(ethnicity, minlength=5) + sex, data=nzis.sub, 
   horizontal=TRUE, las=1)
```
17. Fit a multiple regression with `weekly_income` as the dependent variable,
    and all remaining variables except `random_ID` and `surf_ID` as explanatory
    variables  (hint: `?lm`).  Store the fit as `fit.1`.
```{r}
fit.1 <- lm(weekly_income ~ age_category + sex + ethnicity + 
            highest_qualification + weekly_hours, data=nzis.sub)
```
18. Summarise the fit (hint: `summary`).
```{r}
summary(fit.1)
```
19. Can you apply the model to yourself?
```{r}
# The author is a male between 35 and 59 years of age, with bachelor or higher,
# and who works 37.5 hours per week.  So:
sum(fit.1$coefficients * 
       c(1, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 37.5))
# killing it.
```
20. Call `plot` on `fit.1` and see what happens.
```{r}
plot(fit.1)
```

## Shiny

#### Exercises

1. Extend the app by adding a dropdown box which includes the options:
   normal, uniform, and exponential; and make appropriate changes such that the
   histogram plots data drawn from the selected distribution (hint:
   `?selectInput`, `?rnorm`, `?runif`, and `?rexp`).
```{r}
library(shiny)

ui <- fluidPage(
  titlePanel("Statistical Distributions..."),
  sidebarLayout(
    sidebarPanel(
      selectInput("dist", "select distribution:",
                  choices=c("Normal"="rnorm", "Uniform"="runif", "Exponential"="rexp")),
      numericInput("n", "number of observations:", 
                   value=30, width = "100%"),
      actionButton("go", "Go", icon = icon("refresh"))
    ),
    mainPanel(
      plotOutput("distplot")
    )
  )
)

server <- function(input, output){
  plotdata <- reactive({
    input$go
    n <- isolate(input$n)
    dist <- isolate(input$dist)
    match.fun(dist)(n)
  })
  
  output$distplot <- renderPlot({
    hist(plotdata(), col="grey", border='white', 
         freq=FALSE, xlab="X", main="")
  })
}

shinyApp(ui = ui, server = server)
```
2. If you're feeling particularly adventurous, add additional distributions, as
   well as other inputs required by various distributions.  For example,
   `?rnorm` defaults to the standard normal distribution, but other values of
   $\mu$ and $\sigma$ are possible other than 0 and 1; `?rt` requires a
   degrees of freedom paramater to be present, and so on.  Hint: `?uioutput`.
```{r}
library(shiny)

ui <- fluidPage(
  titlePanel("Statistical Distributions..."),
  sidebarLayout(
    sidebarPanel(
      selectInput("dist", "select distribution:",
                  choices=c("Normal", "Uniform", "Exponential")),
      uiOutput("distpars"),
      numericInput("n", "number of observations:", 
                   value=1000, width = "100%"),
      actionButton("go", "Go", icon = icon("refresh"))
    ),
    mainPanel(
      plotOutput("distplot")
    )
  )
)

server <- function(input, output){
  output$distpars <- renderUI({
    res <- list()
    if (input$dist == "Normal"){
      res[[1]] <- numericInput("mu", "\u03bc:", value = 0)
      res[[2]] <- numericInput("sigma", "\u03c3:", value = 1)
    }
    else if (input$dist == "Uniform"){
      res[[1]] <- numericInput("min", "min:", value = 0)
      res[[2]] <- numericInput("max", "max:", value = 1)
    }
    else if (input$dist == "Exponential"){
      res[[1]] <- numericInput("lambda", "\u03bb:", value = 1)
    }
    res
  })
  
  # a helper function.  return x if x is not null, and y otherwise.
  isnull <- function(x, y){
    if (is.null(x)) y
    else x
  }
  
  plotdata <- reactive({
    input$go
    n <- isolate(input$n)
    dist <- isolate(input$dist)
    if (dist == "Normal"){
      return(rnorm(n, isnull(isolate(input$mu), 0), isnull(isolate(input$sigma), 1)))
    }
    else if (dist == "Uniform"){
      return(runif(n, isnull(isolate(input$min), 0), isnull(isolate(input$max), 1)))
    }
    else if (dist == "Exponential"){
      return(rexp(n, isnull(isolate(input$lambda), 1)))
    }
    NULL
  })
  
  output$distplot <- renderPlot({
    if (is.null(plotdata())) return(NULL)
    hist(plotdata(), col="grey", border='white', 
         freq=FALSE, xlab="X", main="")
  })
}

shinyApp(ui = ui, server = server)
```